Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-14 paper.tex
Pass through abstract

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. The first
class of method in which dramatic improvements have been made is in the
statistically rigorous inference of ARGs under the Sequentially Markov
Coalescent (SMC) model, which are now practical for tens-to-hundreds of
samples. The second class of recently-developed inference methods are based on
approximate models and heuristics which scale to samples sizes roughly three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one major problem is that the ARGs they 
estimate cannot easily be connected with population genetic theory.
Specifically, because the estimated ARGs lack detail about recombination
events, there is currently no way to compute their likelihood
under a population genetics model. In this paper we provide a 
novel backwards-time formulation of the SMC 
and derive a straightforward definition of the likelihood
of the likelihood of a general class of ARG under the SMC. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to other topological inconveniences such as 
polytomies. This opens up many future directions, most excitingly the possibility
of combining the strengths of both inference approaches. Applying
these ideas may allow us to 
make rigorous inference more scalable, and scalable inference 
more rigorous.
\end{abstract}

>>>>

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. The first
class of method in which dramatic progress has been made is in the
statistically rigorous inference of ARGs under the Sequentially Markov
Coalescent (SMC), which are now practical for tens-to-hundreds of samples. 
The second class of recently-developed inference methods are based on
approximate models and heuristics which scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they 
estimate do not have the topological properties required to compute a 
likelihood under models such as the SMC under present-day formulations.
In particular, hueristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a 
novel backwards-time formulation of the SMC 
and derive a straightforward definition 
of the likelihood of a general class of ARG under this model. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies. 
We discuss the possibilities for inference that this opens.
\end{abstract}

% This opens up many future directions, most excitingly the possibility
% of combining the strengths of both inference approaches. Applying
% these ideas may allow us to 
% make rigorous inference more scalable, and scalable inference 
% more rigorous.
